---
title: "Minnesota Timberwolves Data Analysis"
author: "Carter Brunclik"
date: "2025-04-02"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(nbastatR)
library(grid)
library(jpeg)
library(magick)

#Increased buffer size to account for the large amount of data points we process when looking through the nbastat data.
Sys.setenv("VROOM_CONNECTION_SIZE" = 5000000)

```



```{r, include = FALSE}

#Here we instate each season seperately. This is because doing multiple season in one command creates a table so large that R has trouble computing it properly resulting in a vector of zero length.
timber_2016 <- teams_shots(teams = "Minnesota Timberwolves", seasons = 2016)

timber_2017 <- teams_shots(teams = "Minnesota Timberwolves", seasons = 2017)

timber_2018 <- teams_shots(teams = "Minnesota Timberwolves", seasons = 2018)

timber_2019 <- teams_shots(teams = "Minnesota Timberwolves", seasons = 2019)

timber_2020 <- teams_shots(teams = "Minnesota Timberwolves", seasons = 2020)

timber_2021 <- teams_shots(teams = "Minnesota Timberwolves", seasons = 2021)

timber_2022 <- teams_shots(teams = "Minnesota Timberwolves", seasons = 2022)

timber_2023 <- teams_shots(teams = "Minnesota Timberwolves", seasons = 2023)

#This df should have all the seasons binded together and be properly recognized by R.
timber_df <- bind_rows(timber_2016, timber_2017, timber_2018, timber_2019, timber_2020, timber_2021, timber_2022, timber_2023)

```



```{r, include = FALSE}

shot_chart <- timber_df %>%
  mutate(
    shot_type_group = ifelse(typeShot == "3PT Field Goal", "3PT", "2PT"),
    shot_label = case_when(
      !isShotMade ~ "Missed Shot",
      shot_type_group == "3PT" ~ "3PT Made",
      TRUE ~ "2PT Made"
    )
  )

```



```{r, include = FALSE}

#Read in the Timberwolves court
court_img <- image_read("timberwolves_court2_cropped.jpg")

# Rotate 90° counter-clockwise
court_img_rotated <- image_rotate(court_img, 90)

court_grob <- rasterGrob(
  as.raster(court_img_rotated),
  width = unit(1, "npc"),
  height = unit(1, "npc")
)


```

### A High Level View.


- In the 2016-2017 season the Minnesota Timberwolves heavily favored 2PT shots over 3PT shots. 

- Take notice of the 2015-2019 seasons shot count graphics and their abundance of green (2PT shots made)

- We see the trend of a more equal shooting behavior solidify by the 2019-2020 season.

- 2022-2023 season where we see that the Timberwolves now have a much more balanced count of 2PT and 3PT shots.

- The most even seasons in terms of percentage of shot types between 3PT and 2PT shots are 2019-2020 and 2021-2022.

- The bar charts below the shooting graphics make this trend even easier to see at a glance. 

- A lot of 3PT shots are made from the middle or near the middle of the 3PT line.

- 2019-2020 seems to have a larger proportion of 2PT shot scored from near the rim than other seasons.


In the 2016-2017 season we can see that there is much more green visually on the shot charts than others and the bar chart indicates that 80% of points in that season were 2PTers. 

From the 2016 to 2019 seasons we can see that the team has a preference for 2PT shots as for this time a majority of points season to season are from 2PT instances.

2019-2020 marks the beginning of the trend of a more equal 2PT to 3PT ratio. There note that once the ratio hits about 50:50 the amount of 3PTs does not increase. This makes sense, while 3PT shots become more of the norm they will never compete with how easy 2PT shots are to score.

The 2021-2022 season is where we really see something close to true equilibrium in scoring ratio. This can be seen both by the shot chart and the accompanying bar charts.


```{r, echo = FALSE, fig.height = 10, fig.width = 10}

ggplot(shot_chart, aes(x = locationX, y = locationY)) +
  annotation_custom(
    court_grob,
    xmin = -255, xmax = 255,
    ymin = -48, ymax = 876
  ) +
  geom_point(
    aes(color = shot_label, shape = shot_label),
    alpha = 0.5, size = 2
  ) +
  scale_color_manual(
    values = c(
      "3PT Made" = "blue",
      "2PT Made" = "green",
      "Missed Shot" = "red"
    )
  ) +
  scale_shape_manual(
    values = c(
      "3PT Made" = 16,       
      "2PT Made" = 16,       
      "Missed Shot" = 4
    )
  ) +
  coord_fixed(ratio = 1, xlim = c(-250, 250), ylim = c(-50, 450)) +
  labs(
    title = "Minnesota Timberwolves Shot Chart (2016-2022 Seasons)",
    x = NULL, y = NULL,
    color = "Shot Outcome",
    shape = "Shot Outcome"
  ) +
  theme_minimal() +
  facet_wrap(~ slugSeason)

```


```{r, include = FALSE}
#Setup data to easily show shot behavior change.
timber_summary <- timber_df %>%
  mutate(
    shot_type = ifelse(typeShot == "3PT Field Goal", "3PT", "2PT")
  ) %>%
  group_by(slugSeason, shot_type) %>%
  summarize(
    shot_attempts = n(),
    .groups = "drop"
  ) %>%
  group_by(slugSeason) %>%
  mutate(
    total_shots = sum(shot_attempts),
    pct_of_shots = shot_attempts / total_shots
  )

```



```{r, echo = FALSE}

ggplot(timber_summary, aes(x = shot_type, y = pct_of_shots, fill = shot_type)) +
  geom_col(width = 0.6) +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_fill_manual(values = c("2PT" = "green", "3PT" = "blue")) +
  labs(
    title = "2PT vs 3PT Shot Share by Season",
    subtitle = "Minnesota Timberwolves (2016–2022)",
    x = "Shot Type",
    y = "Percent of Shots",
    fill = "Shot Type"
  ) +
  facet_wrap(~ slugSeason) +
  theme_minimal()



```


\newpage


### The Backend



```{r}

#Here we instate each season seperately. This is because doing multiple seasons in one command creates a table so large that R has trouble computing it properly resulting in a vector of zero length.

timber_2016 <- teams_shots(teams = "Minnesota Timberwolves", seasons = 2016)

timber_2017 <- teams_shots(teams = "Minnesota Timberwolves", seasons = 2017)

timber_2018 <- teams_shots(teams = "Minnesota Timberwolves", seasons = 2018)

timber_2019 <- teams_shots(teams = "Minnesota Timberwolves", seasons = 2019)

timber_2020 <- teams_shots(teams = "Minnesota Timberwolves", seasons = 2020)

timber_2021 <- teams_shots(teams = "Minnesota Timberwolves", seasons = 2021)

timber_2022 <- teams_shots(teams = "Minnesota Timberwolves", seasons = 2022)

timber_2023 <- teams_shots(teams = "Minnesota Timberwolves", seasons = 2023)

#This df should have all the seasons binded together and be properly recognized by R. We can troubleshoot by looking at `head(timber_df)` if need be.
timber_df <- bind_rows(timber_2016, timber_2017, timber_2018, timber_2019, timber_2020, timber_2021, timber_2022, timber_2023)

```

---

```{r}

#Here we create a version of our df that focuses on identifying if shots were made or not and their type (2PT or 3PT)

#I collect this information here so I don't have to nest everything in the ggplot.

shot_chart <- timber_df %>%
  mutate(
    shot_type_group = ifelse(typeShot == "3PT Field Goal", "3PT", "2PT"), #ifesle and case_when use simple true false areguements to categorize each shot pulled from timber_df
    shot_label = case_when(
      !isShotMade ~ "Missed Shot",
      shot_type_group == "3PT" ~ "3PT Made",
      TRUE ~ "2PT Made"
    )
  )

```

---

```{r}

#Read in the timberwolves court (the plot backdrops) into an object we can manipulate.
court_img <- image_read("timberwolves_court2_cropped.jpg")

# Rotate 90° counter-clockwise to fit our plot's orientation
court_img_rotated <- image_rotate(court_img, 90)

#Turning the court image into a grid graphical object (grob) so that image size edits can be made in ggplot.
court_grob <- rasterGrob(
  as.raster(court_img_rotated),
  width = unit(1, "npc"),
  height = unit(1, "npc")
)


```

---

```{r, fig.height = 10, fig.width = 10}

ggplot(shot_chart, aes(x = locationX, y = locationY)) +
  
  #Adding the court grob
  annotation_custom(
    court_grob,
    xmin = -255, xmax = 255,
    ymin = -48, ymax = 876
  ) +
  
  #Placing our points, calling back to the shot_chart R chunk for what do to for shape and color.
  geom_point(
    aes(color = shot_label, shape = shot_label),
    alpha = 0.5, size = 2
  ) +
  
  #Giving our points color to make them stand out and make information easier to digest. (Also creates custom legend)
  scale_color_manual(
    values = c(
      "3PT Made" = "blue",
      "2PT Made" = "green",
      "Missed Shot" = "red"
    )
  ) +
  
  #Tells the ggplot what shape to use for each point. (Also helps with legend by giving it shapes)
  scale_shape_manual(
    values = c(
      "3PT Made" = 16,       
      "2PT Made" = 16,       
      "Missed Shot" = 4
    )
  ) +
  
  #Gives our points a grid to plot on.
  coord_fixed(ratio = 1, xlim = c(-250, 250), ylim = c(-50, 450)) +
  labs(
    title = "Minnesota Timberwolves Shot Chart (2016-2022 Seasons)",
    x = NULL, y = NULL,
    color = "Shot Outcome",
    shape = "Shot Outcome"
  ) +
  theme_minimal() +

  #Show multiple seasons on separate plots
  facet_wrap(~ slugSeason)

```

---

```{r}

#Create a new df (similar to before with shot_chart) with the primary focus of finding the shot percentage of 2PTs and 3PTs
timber_summary <- timber_df %>%
  mutate(
    shot_type = ifelse(typeShot == "3PT Field Goal", "3PT", "2PT")
  ) %>%
  group_by(slugSeason, shot_type) %>%
  summarize(
    #Finding shot attempts, the group_by before this has both the number of 2PTs and 3PTs counted.
    shot_attempts = n(),
    .groups = "drop"
  ) %>%
  group_by(slugSeason) %>%
  mutate(
    total_shots = sum(shot_attempts),
    pct_of_shots = shot_attempts / total_shots
  )

```

---

```{r}

#Simple ggplot col charts displaying percentage of 2PT shots and 3PTs shots faceted by season.

ggplot(timber_summary, aes(x = shot_type, y = pct_of_shots, fill = shot_type)) +
  geom_col(width = 0.6) +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_fill_manual(values = c("2PT" = "green", "3PT" = "blue")) +
  labs(
    title = "2PT vs 3PT Shot Share by Season",
    subtitle = "Minnesota Timberwolves (2016–2022)",
    x = "Shot Type",
    y = "Percent of Shots",
    fill = "Shot Type"
  ) +
  facet_wrap(~ slugSeason) +
  theme_minimal()



```


\newpage


### Missing the Point


- In my original plots I included misses but did not comment on them. Here I want to take some time to look at some fun missed shots statistics.

- From the plot I can see that the only top misser that consistently has more makes than misses is Karl-Anthony Towns.

-There is a large disparity of misses and makes for Anthony Edwards.

- Andrew Wiggins holds the spot for the most amount of misses in a season for the Timberwolves (2016-2022) from his time in the 2017 season.

- Gorgui Dieng seemes to have the most even make and miss ratio.



```{r, echo = FALSE}
#Count missed shots per player per season
misses_by_player <- timber_df %>%
  filter(isShotMade == FALSE) %>%
  group_by(yearSeason, namePlayer) %>%
  summarize(
    missed_shots = n(),
    .groups = "drop"
  )

#Get top 6 players with most total missed shots (overall)
top_players <- misses_by_player %>%
  group_by(namePlayer) %>%
  summarize(total_misses = sum(missed_shots)) %>%
  slice_max(total_misses, n = 6)

#Filter and pull names of top missers
filtered_misses <- misses_by_player %>%
  filter(namePlayer %in% top_players$namePlayer)

#Get made shots for same players
made_by_player <- timber_df %>%
  filter(isShotMade == TRUE, namePlayer %in% top_players$namePlayer) %>%
  group_by(yearSeason, namePlayer) %>%
  summarize(made_shots = n(), .groups = "drop")

#Plot
ggplot() +
  
  # Solid lines for missed shots
  geom_line(
    data = filtered_misses,
    aes(x = yearSeason, y = missed_shots, color = namePlayer),
    linewidth = 1
  ) +
  geom_point(
    data = filtered_misses,
    aes(x = yearSeason, y = missed_shots, color = namePlayer),
    size = 2
  ) +
  
  # Dotted lines for made shots
  geom_line(
    data = made_by_player,
    aes(x = yearSeason, y = made_shots, color = namePlayer),
    linetype = "dotted",
    linewidth = 1
  ) +
  geom_point(
    data = made_by_player,
    aes(x = yearSeason, y = made_shots, color = namePlayer),
    shape = 1,
    size = 2
  ) +
  labs(
    title = "Missed vs Made Shots by Top 6 Timberwolves Missers (2016–2022)",
    subtitle = "Solid lines = Misses | Dotted lines = Makes",
    x = "Season",
    y = "Number of Shots",
    color = "Player"
  ) +
  theme_minimal()

```






